LCS Tool: A Computational Platform for Lagrangian Coherent Structures

### 2014-11-07 main.tex
Add bibliography references

More recent work has focused on an exact mathematical formulation of the properties defining LCSs\citep{haller11:_lagran_coher_struc,farazmand12:_comput_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows,haller13:_coher_lagran,blazevski14:_hyper_ellip_trans_barrier_three}. In two-dimensional flows, hyperbolic and parabolic LCSs turn out to be stationary curves of the averaged material shear\citep{farazmand14:_shearless}, whereas elliptic LCSs are stationary curves of the averaged strain\citep{haller13:_coher_lagran}. These variational formulations lead to explicit solutions for LCSs as null-geodesics of appropriate Lorentzian metrics.

@@

\begin{sloppypar}
We seek positions of closed material lines at time $t_0$ that prevail as coherent Lagrangian vortex boundaries (or \emph{elliptic LCSs}) over a time interval $[t_0,t]\subset[t_-,t_+]$. \citet{haller13:_coher_lagran} argue that such initial material line positions are closed stationary curves of the averaged strain functional
\[
Q(\gamma) = \frac{1}{\sigma} \int_0^{\sigma} \frac{\sqrt{\langle r'(s),C_{t_0}^t(r(s))r'(s)\rangle}}{\sqrt{\langle r'(s),r'(s)\rangle}} \dif s,
\]
obtained by averaging the tangential strain arising over $[t_0,t]$ along closed material lines parametrized as $r(s)$ with $s \in [0,\sigma].$
Solutions to this variational problem turn out to be closed orbits of one of two parametrized vector-field families
\begin{equation}
\eta_\pm^\lambda = \sqrt{\frac{\lambda_2 - \lambda^2}{\lambda_2 - \lambda_1}} \xi_1 \pm \sqrt{\frac{\lambda^2 - \lambda_1}{\lambda_2 - \lambda_1}}\xi_2,
\label{eq:eta}
\end{equation}
with $\lambda > 0$ playing the role of a parameter. Such closed orbits satisfy the differential equation
\begin{equation}
r' = \eta_\pm^\lambda(r),
\label{eq:etafields}
\end{equation}
which coincide with null-geodesics of the Lorentzian metric family 
\[
e_\lambda(u,v) = \left\langle u,\left[D_{t_0}^t(r) - \lambda^2 I\right] v \right\rangle.
\]
For this reason, we refer to the computation of elliptic LCSs as limit cycles of \cref{eq:etafields} as \emph{geodesic detection of elliptic LCSs}.
\end{sloppypar}

Any orbit of \cref{eq:etafields} turns out to stretch uniformly under the flow map $F_{t_0}^t$. Specifically, any subset of an orbit of \cref{eq:etafields} increases its arc length precisely by a factor of $\lambda$. For this reason, we refer to trajectories of \cref{eq:etafields} as $\lambda$-lines. Following \citet{haller13:_coher_lagran}, we call the outermost member of a closed family of $\lambda$-lines a coherent Lagrangian vortex boundary.

@@

As discussed in \cref{sec:Elliptic LCSs}, positions of elliptic LCSs at time $t_0$ are found as closed orbits of the $\eta_\pm^\lambda$ vector fields defined in \cref{eq:eta}. We find such orbits by integrating \cref{eq:etafields} from points of an appropriately chosen section (Poincare section), and evaluating the first return map (Poincare map) onto this section. A $\lambda$-line returning to its starting point is then an elliptic LCS. The outermost member of a family of closed $\lambda$-lines (obtained by varying $\lambda$) is a coherent Lagrangian vortex boundary\citep{haller13:_coher_lagran}.

@@

The final example demonstrates the use of LCS Tool on velocity data derived from satellite-observed sea-surface heights under the geostrophic approximation. In contrast to the previous two analytic examples, the velocity field is available only with discrete temporal and spatial resolution. Our region of interest is a small domain in the South Atlantic Ocean, where exceptionally coherent eddies, Agulhas rings, were recently found by \citet{haller13:_coher_lagran} using the theory we surveyed in \cref{sec:Elliptic LCSs}.

@@

\Cref{f:ocean dataset LCS} shows the positions of elliptic and hyperbolic LCSs on 22 November 2006, which is the same time instant analysed in \citet{haller13:_coher_lagran,beron-vera13:_objec_agulh}. Here the integration time is chosen as $T = 30$ days, as opposed to 90 days in the cited references, to avoid tangling of hyperbolic LCSs. Our analysis via LCS Tool reveals five coherent eddy boundaries. The large coherent eddy at $(3,-32)$ has a non-stretching boundary, i.e., $\lambda = 1$. It corresponds to eddy \#2 in Figure~3 of \citet{beron-vera13:_objec_agulh}. Additionally, four smaller coherent eddy cores are found. They do not stay coherent over a time of 90 days. All closed orbits are stable under increased spatial resolution for the Cauchy-Green strain tensor field. Repelling LCSs (red) and attracting LCSs (blue) are superimposed over the plot of FTLE values. The hyperbolic LCSs determine the deformation of the fluid in-between coherent eddy cores.

>>>>

More recent work has focused on an exact mathematical formulation of the properties defining LCSs\citep{haller11:_lagran_coher_struc,farazmand12:_comput_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows,haller13:_coher_lagran,haller14:_adden_coher_lagran,farazmand13:_attrac_lagran,blazevski14:_hyper_ellip_trans_barrier_three}.
In two-dimensional flows, hyperbolic and parabolic LCSs turn out to be stationary curves of the averaged material shear\citep{farazmand14:_shearless}, whereas elliptic LCSs are stationary curves of the averaged strain\citep{haller13:_coher_lagran,haller14:_adden_coher_lagran}.
These variational formulations lead to explicit solutions for LCSs as null-geodesics of appropriate Lorentzian metrics.

@@

\begin{sloppypar}
We seek positions of closed material lines at time $t_0$ that prevail as coherent Lagrangian vortex boundaries (or \emph{elliptic LCSs}) over a time interval $[t_0,t]\subset[t_-,t_+]$. \citet{haller13:_coher_lagran,haller14:_adden_coher_lagran} argue that such initial material line positions are closed stationary curves of the averaged strain functional
\[
Q(\gamma) = \frac{1}{\sigma} \int_0^{\sigma} \frac{\sqrt{\langle r'(s),C_{t_0}^t(r(s))r'(s)\rangle}}{\sqrt{\langle r'(s),r'(s)\rangle}} \dif s,
\]
obtained by averaging the tangential strain arising over $[t_0,t]$ along closed material lines parametrized as $r(s)$ with $s \in [0,\sigma].$
Solutions to this variational problem turn out to be closed orbits of one of two parametrized vector-field families
\begin{equation}
\eta_\pm^\lambda = \sqrt{\frac{\lambda_2 - \lambda^2}{\lambda_2 - \lambda_1}} \xi_1 \pm \sqrt{\frac{\lambda^2 - \lambda_1}{\lambda_2 - \lambda_1}}\xi_2,
\label{eq:eta}
\end{equation}
with $\lambda > 0$ playing the role of a parameter. Such closed orbits satisfy the differential equation
\begin{equation}
r' = \eta_\pm^\lambda(r),
\label{eq:etafields}
\end{equation}
which coincide with null-geodesics of the Lorentzian metric family 
\[
e_\lambda(u,v) = \left\langle u,\left[D_{t_0}^t(r) - \lambda^2 I\right] v \right\rangle.
\]
For this reason, we refer to the computation of elliptic LCSs as limit cycles of \cref{eq:etafields} as \emph{geodesic detection of elliptic LCSs}.
\end{sloppypar}

Any orbit of \cref{eq:etafields} turns out to stretch uniformly under the flow map $F_{t_0}^t$. Specifically, any subset of an orbit of \cref{eq:etafields} increases its arc length precisely by a factor of $\lambda$. For this reason, we refer to trajectories of \cref{eq:etafields} as $\lambda$-lines. Following \citet{haller13:_coher_lagran,haller14:_adden_coher_lagran}, we call the outermost member of a closed family of $\lambda$-lines a coherent Lagrangian vortex boundary.

@@

As discussed in \cref{sec:Elliptic LCSs}, positions of elliptic LCSs at time $t_0$ are found as closed orbits of the $\eta_\pm^\lambda$ vector fields defined in \cref{eq:eta}. We find such orbits by integrating \cref{eq:etafields} from points of an appropriately chosen section (Poincare section), and evaluating the first return map (Poincare map) onto this section. A $\lambda$-line returning to its starting point is then an elliptic LCS. The outermost member of a family of closed $\lambda$-lines (obtained by varying $\lambda$) is a coherent Lagrangian vortex boundary\citep{haller13:_coher_lagran,haller14:_adden_coher_lagran}.

@@

The final example demonstrates the use of LCS Tool on velocity data derived from satellite-observed sea-surface heights under the geostrophic approximation. In contrast to the previous two analytic examples, the velocity field is available only with discrete temporal and spatial resolution. Our region of interest is a small domain in the South Atlantic Ocean, where exceptionally coherent eddies, Agulhas rings, were recently found by \citet{haller13:_coher_lagran,haller14:_adden_coher_lagran} using the theory we surveyed in \cref{sec:Elliptic LCSs}.

@@

\Cref{f:ocean dataset LCS} shows the positions of elliptic and hyperbolic LCSs on 22 November 2006, which is the same time instant analysed in \citet{haller13:_coher_lagran,haller14:_adden_coher_lagran,beron-vera13:_objec_agulh}. Here the integration time is chosen as $T = 30$ days, as opposed to 90 days in the cited references, to avoid tangling of hyperbolic LCSs. Our analysis via LCS Tool reveals five coherent eddy boundaries. The large coherent eddy at $(3,-32)$ has a non-stretching boundary, i.e., $\lambda = 1$. It corresponds to eddy \#2 in Figure~3 of \citet{beron-vera13:_objec_agulh}. Additionally, four smaller coherent eddy cores are found. They do not stay coherent over a time of 90 days. All closed orbits are stable under increased spatial resolution for the Cauchy-Green strain tensor field. Repelling LCSs (red) and attracting LCSs (blue) are superimposed over the plot of FTLE values. The hyperbolic LCSs determine the deformation of the fluid in-between coherent eddy cores.
